Red Wine Quality by Abhishek Patria
========================================================



```{r echo=FALSE, warning=FALSE, message=FALSE, warning=FALSE, packages}

library(ggplot2)
library(dplyr)
library(GGally)
library(tidyr)
library(gridExtra)

```

```{r global_options, include=FALSE}

knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE)

```


```{r echo=FALSE, warning=FALSE, Load_the_Data}

#setwd("'''Your working directory path goes here.'''")

df <- read.csv("wineQualityReds.csv")

```

# Description
>In the given project, I will explore the data set on red wines' quality. The 
primary motive is to explore the influence of chemical properties on the quality
of red wines. This data set contains 1,599 red wines with 11 variables. The data
set is available  [here](https://s3.amazonaws.com/udacity-hosted-downloads/ud651/wineQualityReds.csv) and information about the data set is available [here](https://s3.amazonaws.com/udacity-hosted-downloads/ud651/wineQualityInfo.txt).


# Dataset Quick-Summary

#### Dataset dimentions:

```{r echo=FALSE,warning=FALSE, Dimentions}
dim(df)
```

#### Dataset structure:
```{r echo=FALSE, warning=FALSE, Structure}
str(df)
```


#### Dataset summary:
```{r echo=FALSE, warning=FALSE, Summary}
summary(df)
```
>If we see the quality variable, it is an integer variable with a range of 
[3,8] having a median quality of 6 with the average of 5.636 which might imply a
slightly left skewed distribution.


```{r echo=FALSE, warning=FALSE, Quality}

ggplot(df,aes(as.factor(quality))) + 
  geom_bar() + xlab("Quality") + ylab("Count") 

```

>We can see from the barplot that the distribution is almost symmetrical 
about its mean value of 5.636 and a slight left skewness.


# Pairwise Matrix

```{r echo=FALSE, warning=FALSE, Pairwise}

ggpairs(data = df, 
  lower = list(
    continuous = wrap("smooth_loess",alpha = 0.1, color = "red", size=0.5) 
    ),
  upper = list(continuous = wrap(ggally_cor, size = 2, color = "black"))
  ) +
theme(
    axis.text = element_text(size = 6),
    axis.title = element_text(size = 6),
    legend.background = element_rect(fill = "white"),
    panel.grid.major = element_line(colour = NA),
    panel.grid.minor = element_blank(),
    panel.background = element_rect(fill = "grey95")
    
)

```

# Univariate Plots

#### Alcohol Content

```{r echo=FALSE, warning=FALSE, uni_functions}

uni_function <- function(x, xlab){
  
  ggplot(df,aes(x)) + 
    geom_histogram(aes(fill=..count..)) + xlab(xlab) + ylab("Count") +
    scale_fill_gradient(low="pink", high="dark red")

  }

uni_function_log10 <- function(x, xlab){
  
  ggplot(df,aes(x)) + 
    geom_histogram(aes(fill=..count..)) + xlab(paste(xlab, " : log10", sep="")) + ylab("Count") +
    scale_fill_gradient(low="pink", high="dark red") + scale_x_log10()

  }

uni_function_sqrt <- function(x, xlab){
  
  ggplot(df,aes(x)) + 
    geom_histogram(aes(fill=..count..)) + xlab(paste(xlab, " : sqrt", sep="")) + ylab("Count") +
    scale_fill_gradient(low="pink", high="dark red") + scale_x_sqrt()

  }

```


```{r echo=FALSE, warning=FALSE, Alcohol_Content}

p1 <- uni_function(df$alcohol, "Alcohol")
p2 <- uni_function_log10(df$alcohol, "Alcohol")
p3 <- uni_function_sqrt(df$alcohol, "Alcohol")

grid.arrange(p1, p2, p3, ncol=2)

```

>Being a tailed distribution, plotting the alcohol content with a log and sqrt 
scale does not help us understanding it.

#### Residual Sugar

```{r echo=FALSE, warning=FALSE, Residual_Sugar}

p1 <- uni_function(df$residual.sugar, "Residual Sugar")
p2 <- uni_function_log10(df$residual.sugar, "Residual Sugar")
p3 <- uni_function_sqrt(df$residual.sugar, "Residual Sugar")

grid.arrange(p1, p2, p3, ncol=2)

```

>Residual Sugar is the content that provides with sweet after taste after you
take a wine sip. Also, it is an indicator of the quality of fermentation process.
Low residual sugar may mean better fermentation process and thus higher alcohol
content. It would be interesting to take a look at this in bivariate analysis.


#### Free sulphur, Total sulfur, Bound sulfur

>I have created a new variable "Bound sulfurdioxide" which is Total sulfurdioxide
less the free sulfur dioxide, as the free sulfurdioxide has a lot of dispersed
distribution.

```{r echo=FALSE, warning=FALSE, sulfur}


p1 <- uni_function(df$free.sulfur.dioxide, "Free sulfurdioxide")
p2 <- uni_function(df$total.sulfur.dioxide, "Total sulfurdioxide")
p3 <- uni_function((df$total.sulfur.dioxide - df$free.sulfur.dioxide) , "Bound sulfurdioxide")
  
grid.arrange(p1,p2,p3, ncol=2)


```

>When plotting on log scale the curve starts acting normally distributed
 
```{r echo=FALSE, warning=FALSE, sulfur_log}

p1 <- uni_function_log10(df$free.sulfur.dioxide, "Free sulfurdioxide : log10")
p2 <- uni_function_log10(df$total.sulfur.dioxide, "Total sulfurdioxide : log10")
p3 <- uni_function_log10((df$total.sulfur.dioxide - df$free.sulfur.dioxide) , "Bound sulfurdioxide : log10")
 
grid.arrange(p1,p2,p3, ncol=2)

df$bound.sulfur.dioxide <-  with(df, total.sulfur.dioxide - free.sulfur.dioxide)

```

#### Sulphates
```{r echo=FALSE, warning=FALSE, sulphates}


p1 <- uni_function(df$sulphates, "Sulphates")
p2 <- uni_function_log10(df$sulphates, "Sulphates")
p3 <- uni_function_sqrt(df$sulphates, "Sulphates")

grid.arrange(p1,p2,p3, ncol=2)


```

>We see that after applying log scale the distribution tends to be normal but a
lot of outliers still remain.


#### Citric acid
```{r echo=FALSE, warning=FALSE, citric_acid}


p1 <- uni_function(df$citric.acid, "Citric Acid")
p2 <- uni_function_log10(df$citric.acid, "Citric Acid")
p3 <- uni_function_sqrt(df$citric.acid, "Citric Acid")

grid.arrange(p1,p2,p3, ncol=2)


```

>We see that plotting citric acid on log scale flips the curve on normal scale,
while plotting it on sqrt scale starts giving it normal featrue but with a couple
of extreme outliers.

#### Fixed acidity, Volatile acidity, Total acidity

>I have created a new variable "Bound sulfurdioxide" which is Total sulfurdioxide
less the free sulfur dioxide, as the free sulfurdioxide has a lot of dispersed
distribution.

```{r echo=FALSE, warning=FALSE, acidity}


p1 <- uni_function(df$fixed.acidity, "Fixed Acidity : log10")
p2 <- uni_function(df$volatile.acidity, "Volatile Acidity : log10")
p3 <- uni_function((df$fixed.acidity + df$volatile.acidity) , "Total Acidity : log10")
  
grid.arrange(p1,p2,p3, ncol=2)

df$total.acidity <-  with(df, fixed.acidity + volatile.acidity)

```

>When plotting on log scale the curve starts acting normally distributed
 
```{r echo=FALSE, warning=FALSE, acidity_log}

p1 <- uni_function(df$fixed.acidity, "Fixed Acidity")
p2 <- uni_function(df$volatile.acidity, "Volatile Acidity")
p3 <- uni_function((df$fixed.acidity + df$volatile.acidity) , "Total Acidity")

grid.arrange(p1,p2,p3, ncol=2)

```

#### Density

```{r echo=FALSE, warning=FALSE, Density}

p1 <- uni_function(df$density, "Density")
p2 <- uni_function_log10(df$density, "Density")
p3 <- uni_function_sqrt(df$density, "Density")

grid.arrange(p1,p2,p3, ncol=2)

```



# Univariate Analysis

### Dataset structure:

* We can see that the dataset has 13 variables with 1599 observations
* All the data is of numbers type with integer and numeric being the subtypes
* The variable "x" acts as a unique ID for each row of the data

### Points of interest: 
1. pH and density are almost normally distributed
2. Most wines have an alcohol content of less than 11%
3. Residual sugar, chlorides have extreme outliers
4. Sulfur dioxide seems to have normal distribution on a log scale


### Main feature of interest:
It is to see how the quality is being affected by changes in other variables

### Other features in the dataset that will help support my investigation:
I think that Residual sugar, alcohol, sufur, acidity and pH will be the main
features that will affect the wine taste. Density may also contribute to the 
smoothness of the wine.


### New variables created

I have created two new variables as follows:-

1. Bound sulfurdioxide = Total sulfurdioxide - Free sulfurdioxide
2. Total acidity = Fixed acidity + Volatile acidity

Though, I was not able to derive insights from their univariate analysis.


### Of the features you investigated, were there any unusual distributions? \
Did you perform any operations on the data to tidy, adjust, or change the form \
of the data? If so, why did you do this?

I had a feeling that most of the data would be normally distributed but then 
analyzing the variables came as a shocker. This provided a sense how data is in
real life.

The data had extreme outliers in the case of residual sugar, chlorides, sulphates.
This might represent a case of product differentiation among the wines tested.


# Bivariate Plots Section


```{r echo=FALSE, warning=FALSE, Bivariate_Plots}

corr_mat_data <- subset(df, select=-c(1))

ggcorr(corr_mat_data, nbreaks = 8, low="dark red", high="dark red", label=TRUE, 
       label_size = 3, size = 2, label_round = 2)

```

>From the correlation matrix we can see that quality has the strongest relation
with alcohol, volatile acidity, sulphates and citric acid with the values being 
0.5, -0.4, 0.3, and -0.2 respectively: while correlation value of 0 with residual
sugar implies that residual sugar does not contribute to quality.


###Let us first plot quality against various variables

```{r echo=FALSE, warning=FALSE, bi_function}

bi_function <- function(x,y,ylab){
  
  ggplot(aes(x=factor(x), y=y), data = df) +
  geom_boxplot( alpha = .3,color = 'dark green') +
  geom_point(alpha = .2, position = "jitter", col="dark red") +
  stat_summary(fun.y = "median", geom = "point", color = "purple", shape = 18, size = 3) +
  geom_smooth(aes(quality-2,y ), method = "lm", se = FALSE, size=1, color = 'black')+
  xlab("Quality") + ylab(ylab)
  
}

```


>Alcohol vs Quality

```{r echo=FALSE, warning=FALSE, bi_alcohol}

bi_function(df$quality,df$alcohol,"Alcohol")
cor.test(df$alcohol, df$quality)

```


>Volatile Acidity vs Quality

```{r echo=FALSE, warning=FALSE, bi_volatileacidity}

bi_function(df$quality,df$volatile.acidity,"Volatile Acidity")
cor.test(df$volatile.acidity, df$quality)

```


>Sulphates vs Quality

```{r echo=FALSE, warning=FALSE, bi_sulphates}

bi_function(df$quality,df$sulphates,"Sulphates")
cor.test(df$sulphates, df$quality)

```


>Citric Acid vs Quality

```{r echo=FALSE, warning=FALSE, bi_citricacid}

bi_function(df$quality,df$citric.acid,"Citric Acid")
cor.test(df$citric.acid, df$quality)

```


>Total Sulfhur Dixoide vs Quality

```{r echo=FALSE, warning=FALSE, bi_totalsulfur_dioxide}

bi_function(df$quality,df$total.sulfur.dioxide,"Total Sulfur Dioxide")
cor.test(df$total.sulfur.dioxide, df$quality)

```


>pH vs Quality

```{r echo=FALSE, warning=FALSE, bi_pH}

bi_function(df$quality,df$pH,"pH")
cor.test(df$pH, df$quality)

```


>Density vs Quality

```{r echo=FALSE, warning=FALSE, bi_density}

bi_function(df$quality,df$density,"Density")
cor.test(df$density, df$quality)

```


>Free Sulfur Dioxide vs Quality

```{r echo=FALSE, warning=FALSE, bi_freesulfurdioxide}

bi_function(df$quality,df$free.sulfur.dioxide,"Free Sulfur Dioxide")
cor.test(df$free.sulfur.dioxide, df$quality)

```


>Chlorides vs Quality

```{r echo=FALSE, warning=FALSE, bi_chlorides}

bi_function(df$quality,df$chlorides,"Chlorides")
cor.test(df$chlorides, df$quality)

```


>Residual Sugar vs Quality

```{r echo=FALSE, warning=FALSE, bi_residualsugar}

bi_function(df$quality,df$residual.sugar,"Residual Sugar")
cor.test(df$residual.sugar, df$quality)

```


>Fixed Acidity vs Quality

```{r echo=FALSE, warning=FALSE, bi_fixedacidity}

bi_function(df$quality,df$fixed.acidity,"Fixed Acidity")
cor.test(df$fixed.acidity, df$quality)

```


>Now let us try to plot fixed acidity with quality and density on a log scale 
as we see from the correlation matrix that they have a strong correlation
number.


>Fixed Acidity vs Density

```{r echo=FALSE, warning=FALSE, bi_fixedacidity_density}

ggplot(df, aes(x=fixed.acidity, y=density)) +
  geom_smooth(method="loess", col="dark green") +
  geom_point(alpha = 0.2 , position = "jitter" , col = "dark red") +
  xlab("Density") + ylab("Fixed Acidity") + scale_x_log10()


cor.test(df$density, log(df$fixed.acidity))

```


>Fixed Acidity vs pH

```{r echo=FALSE, warning=FALSE, bi_fixedacidity_pH}

ggplot(df, aes(x=fixed.acidity, y=pH)) +
  geom_smooth(method="loess", col="dark green") +
  geom_point(alpha = 0.2 , position = "jitter" , col = "dark red") +
  xlab("pH") + ylab("Fixed Acidity") + scale_x_log10()


cor.test(df$pH, log(df$fixed.acidity))

```

>The smoothened curve can be taken as a near straight line.


# Bivariate Analysis


### Quality against other variables
>I was not able to find a strong correlation between quality and other variables.
But if seen as best amongst bad, alcohol, volatile acidity, sulphates and citric
acid are the variables that may be used in developing a linear model to predict 
quality.

>Another thing to notice among the four selected variables, only volatile acidity 
is negatively correlated. Also, no satisfactory dependence could be drawn from
the t-tests conducted on various pairs of variables with quality.



### Relationships between variables which are not the main feature of interest
>On analysing fixed acidity with the log scale plots  of density and pH 
respectively, I found a strong correlation of fixed density with the other two. 
This can be supported from the line graph plotted and the confidence
interval band around it.


### What was the strongest relationship you found?
>With respect to quality, alcohol had the strongest relationship while if we say
about fixed acidity then log(density) had the strongest one.



# Multivariate Plots Section


```{r echo=FALSE, warning=FALSE, scatter_breakup}

scatter_breakup <- function(x, xlab, y, ylab, breakup, breakup_lab) {
  ggplot(data=df, aes(x=x, y=y, color=factor(breakup)) ) +
   geom_jitter(size=1) +
   geom_smooth(method = "lm", se=FALSE, size = 0.5) +
   scale_color_brewer(type='qual', guide=guide_legend(title=breakup_lab),
                         palette = 'Reds') +
   xlab(xlab) + ylab(ylab)
}


```


>I have plotted fixed acidity against density and pH with quality breakup as 
these combinations have the strongest correlation in the correlation matrix.

```{r echo=FALSE, warning=FALSE, fixed_acidity_vs_density}  

scatter_breakup(df$fixed.acidity,"Fixed Acidity  (gm/dm^3)", 
                df$density,"Density  (gm/dm^3)", 
                df$quality, "Quality")

```


```{r echo=FALSE, warning=FALSE, fixed_acidity_vs_pH}

scatter_breakup(df$fixed.acidity,"Fixed Acidity  (gm/dm^3)", df$pH,"pH", 
                df$quality, "Quality")

```

>With alcohol and volatile acidity being strongly correlated with the quality of
the red wine, it is a must to explore the dependence of these two with their 
respective strongest correlations. As a result I have plotted density vs alcohol
and citric acid vs volatility.

```{r echo=FALSE, warning=FALSE, density_vs_alcohol}

scatter_breakup(df$density,"Density  (gm/dm^3)", df$alcohol,"Alcohol (% volume)", 
                df$quality, "Quality")

```


```{r echo=FALSE, warning=FALSE, citric_acid_vs_volatile_acidity}

scatter_breakup(df$citric.acid,"Citric Acid  (gm/dm^3)", 
                df$volatile.acidity,"Volatile Acidity", 
                df$quality, "Quality")

```

# Multivariate Analysis

### Relationships observed:
Density and fixed acidity go hand in hand and they are apparently converging
at higher units.

There is a clear cut interdependence of alcohol on density. For a given alcohol
level, higher density means higher quality.

### Surprising interactions between features:

Though there is a strong correlation between pH and fixed acidity, this does not
reveal any individual insight about the quality as around an acidity of 6.5 the
trend changes for quality.


------

# Final Plots and Summary

### Plot One
```{r echo=FALSE, warning=FALSE, Plot_Three}

ggplot(df,aes(as.factor(quality))) + 
  geom_bar() + xlab("Quality") + ylab("Count") +
  ggtitle("Distribution of Quality")
summary(df$quality)

```

### Description One
Wines with quality 5-6 make up most of the space i.e. they are mainstream, while
high or low quality wines have a very small ratio: this in combination provides 
us with the apparently normal distribution of quality of wines.


### Plot Two
```{r echo=FALSE, warning=FALSE, Plot_One}

bi_function(df$quality,df$alcohol,"Alcohol (% volume)") + ggtitle("Alcohol vs Quality")

```

### Description Two
Alcohol has the strongest correlation with the quality when compared to other
variables, it is better to see its pattern with varying quality. From the above 
plot, it can be seen that higher % of alcohol yields better wine.

### Plot Three
```{r echo=FALSE, warning=FALSE, Plot_Two}

p1 <- bi_function(df$quality,df$fixed.acidity,"Fixed Acidity (gm/dm^3)") +
      ggtitle("Fixed Acidity vs Quality")
p2 <- bi_function(df$quality,df$volatile.acidity,"Volatile Acidity (gm/dm^3)") +
      ggtitle("Volatile Acidity vs Quality")
p3 <- bi_function(df$quality,df$total.acidity,"Total Acidity (gm/dm^3)") +
      ggtitle("Total Acidity vs Quality")
p4 <- bi_function(df$quality,df$citric.acid,"Citric Acid (gm/dm^3)") +
      ggtitle("Citric Acid vs Quality")
grid.arrange(p1, p2, p3, p4, ncol = 2)

```

### Description Three
Volatile acid has a strong negative correlation with wine quality fixed acidity, 
total acidity and citric acid has a positive relationship with wine quality. 



------

# Reflection


While the dataset is small in size, it has a lot of variables to explore. The 
analysis performed above is purely statistical and does not include the domain 
knowledge of wine tasting. The dependence of quality can be mainly explained by
alcohol, acid and sulphates.

Also, correlation does not equate to causation. Moreover, domain knowledge is of
utmost importance to clean, understand, generate new insights and inferences 
from the data.